Independent discovery of a nulling pulsar with unusual sub-pulse drifting properties with the Murchison Widefield Array

### 2022-04-28 paper/main.tex
Fixed up some typography issues with 'sim'

We report the independent discovery of PSR \psr{} with the Murchison Widefield Array (MWA) in the ongoing Southern-sky MWA Rapid Two-meter (SMART) pulsar survey.
\psr{} has a period of $\sim1.306\,$s, a dispersion measure (DM) of $\sim\psrdm\,$\dmunit, and a nulling fraction of $\sim$ \nullfraction{}.
This pulsar highlights the advantages of the survey's long dwell times ($\sim80\,$min), which, when fully searched, will be sensitive to the expected population of similarly bright, intermittent pulsars with long nulls.
A single-pulse analysis in the MWA's 140-170 MHz band also reveals a complex sub-pulse drifting behavior, including both rapid changes of the drift rate characteristic of mode switching pulsars, as well as a slow, consistent evolution of the drift rate within modes. In some longer drift sequences, interruptions in the otherwise smooth drift rate evolution occur preferentially at a particular phase, typically lasting a few pulses.
These properties make this pulsar an ideal test bed for prevailing models of drifting behavior such as the carousel model.

@@

Each observation was individually calibrated using {\bf observations of 3C444 taken within $\sim 2$ hours of the respective target observations.
The calibration solutions were obtained using} the Real Time System software \citep{Mitchell2008}.
After calibration, the voltages were used to form a tied-array beam in the direction of the pulsar (the localization procedure is discussed in \S\ref{sec:localization}), using the bespoke beamforming software described in \citet{Ord2019}.
{\bf The MWA is located in a very radio quiet location, and no significant radio frequency interference was found in the beamformed data.}
After determining the best dispersion measure (DM) and period (see \S\ref{sec:period}), de-dispersion and folding were performed with {\tt DSPSR} \citep{VanStraten2011b} and {\tt PSRCHIVE} \citep{Hotan2004} to form single pulse archives, from which Stokes I pulse stacks were created.
Profiles of each detection, as well as an integrated profile incorporating the data from all four observations, is shown in Fig \ref{fig:profiles}.
Subsequent analysis of the pulse stacks is described in \S\ref{sec:analysis}.
\begin{figure}[!th]
    \centering
    \includegraphics[width=\linewidth]{profiles.png}
    \caption{{\bf Peak-normalised mean profiles (black) of four of the observations}, and the total mean profile (top panel) including all four observations, with a time resolution of $\sim 1.3\,$ms. {\bf The two observations on MJD 59002 are not included here as they consist almost entirely of noise.} The profiles show a barely resolved double peak structure. {\bf The gray lines show the profiles made up only of pulses classified as nulls.}}
    \label{fig:profiles}
\end{figure}

@@

Follow-up of the initial MWA detection with the full 80 minutes, as well as in other archival MWA observations, revealed that the pulsar has a very large nulling fraction ($\sim$ \nullfraction{})\footnote{The nulling fraction estimate comes only from analysis of the MWA observations.}, with some nulling sequences {\bf (excluding occasional intermittent pulses) potentially exceeding $20\,$min (i.e. the length of the first observation on MJD 59002, which contained \textit{only} a few intermittent pulses, as shown in Fig. \ref{fig:maxima})}.
A full analysis of the nulling properties is presented in Section \S\ref{sec:nulling} below.
Despite the fact that \psr{} was serendipitously discovered in the first ten minutes, it may well be the harbinger for a population of relatively bright, intermittent pulsars which the SMART survey, with its $80$-minute dwell times, is well-suited to detect.
A similar strategy is employed in the Low Frequency Array (LOFAR) Tied-Array All-Sky Survey \citep[LOTAAS][]{Sanidas2019} and the low latitude ($|b| \le 5^\circ$) segment of the High Time Resolution Universe survey \citep[HTRU][]{Keith2010}, which use dwell times of approximately one hour.

@@

The discovery observation (Obs ID 1226062160) was taken when the MWA was in its \textit{compact} configuration \citep[see][]{Wayth2018}, giving an effective tied-array beam size of $\sim$ 23$^\prime$ at the central frequency of $155\,$MHz.
As per the SMART survey design, these beams are tiled across the field of view, each of which is searched for periodic candidates.
The compact configuration includes many redundant baselines, and the tied-array beam for the compact configuration is complex.
Consequently, the pulsar was detected in multiple adjacent beams, including a boresight beam that yielded the highest signal-to-noise, and several nearby grating lobes.
The ability to exploit grating lobe detections for fast-tracking candidate confirmation and localization will be discussed in the survey description paper \citetext{Bhat et al. in prep}.

The other {\bf five observations} were taken when the MWA was in its extended configuration, giving a tied-array beam size of $\sim$ 3$^\prime$ at $155\,$MHz.
Localization within {\bf the four observations in which the pulsar was bright enough} followed the identical procedure as used in \citet{Swainston2021}, i.e.\ gridding the candidate positions with overlapping tied-array beams and using signal-to-noise and the estimated beam shape to produce a likelihood map of the source position.
It is known that the ionosphere is capable of shifting the apparent position of sources by an appreciable fraction ($\sim10$-$30\%$) of the tied-array beam, and that these offsets are not necessarily corrected by the calibration process when the calibration observations are separated from the target observation, both spatially and temporally \citetext{Swainston et al., submitted}.
Nevertheless, the localization afforded by the extended array observations was sufficiently precise (a few arcminutes) to warrant follow-up with a high-resolution imaging telescope.
Accordingly, images made with Band 3 ($300$-$500\,$MHz) data from the upgraded Giant Metrewave Radio Telescope \citep[uGMRT;][]{Gupta2017}, obtained under Director's Discretionary Time (DDTC185), enabled us to confirm the pulsar's position to an accuracy of $\sim1^\prime$.
The right ascension and declination of the final best position, consistent with both the uGMRT imaging and the MWA extended configuration observations, are 00h26m37s and $-19^\circ$56'28''.

@@

Lacking a timing solution, we used PSRCHIVE's \texttt{pdmp} routine to determine the best period ($P$) and DM for each observation independently. {\bf This routine performs a brute-force grid search in both period and DM parameter space, and returns the period and DM that yields the highest profile signal-to-noise}. The weighted averages {\bf of both period and DM (respectively)} over the four observations were calculated, {\bf with the uncertainties reported by \texttt{pdmp} being} used to weight the individual measurements.
The values thus derived were $P = 1.306150 \pm 0.000005\,$s and $\text{DM} = \psrdm \pm \psrdmerr\,$\dmunit.
This simple estimate does not take the period derivative into account (in fact, it implicitly assumes $\dot{P} = 0$), and we note that a sufficiently large $\dot{P}$ would cause the period to drift significantly over the time period spanned by our four MWA observations ($\sim$ 1.5\,yr).
However, the pulse stacks (e.g.\ Fig. \ref{fig:pulsestacks}) folded on the above period showed no detectable slope, which we estimate would be discernible if the pulse window changed by more than $20\,$ms over the course of the observation.
Given this, we place a conservative upper limit of $|\dot{P}| \le 10^{-13}$\,s\,s$^{-1}$.

@@

\begin{figure*}[t]
    \includegraphics[width=\textwidth]{maxima_plot.png}
    \caption{Peak flux density within the pulse window as a function of time for six MWA observations (MJDs given along the $y$-axis). {\bf The shaded gray regions show detected subpulses that are included in the nulling fraction estimation, as explained in the text.} The peak fluxes were measured after smoothing the time series with a Gaussian filter ($\sim$ 2\,ms wide, the approximate width of a sub-pulse) to suppress the noise contribution and accentuate the contrast between nulling and burst sequences. {\bf Note that the low-level peaks in the $\sim70$-$80\,$mins time range of the MJD 58434 and in the second MJD 59002 observation} are single pulses faintly visible in the pulse stack, but which are faint because the pulsar was moving out of the MWA's primary beam.}
    \label{fig:maxima}
\end{figure*}

To estimate the nulling fraction, we first identified which pulses contained sub-pulses within the pulse window (this is a semi-automated procedure described in \S\ref{sec:drifting_analysis}).
Any pulse without a sub-pulse is then counted as a null, even if it is nested within a burst sequence.
{\bf Under this definition, we report a nulling fraction of $\sim$\nullfraction{}.}

@@

The error on this estimate, however, will be dominated by the small number of burst/null sequences captured in these six observations.
Therefore, the true nulling fraction may ultimately prove to be much lower or higher than the $\sim$\nullfraction{} quoted here.
Ongoing follow-up observations (and a more complete search through archival MWA data) will allow us to place much stronger constraints on the nulling fraction, and will be reported in a future publication.

@@

The distinction between the subcategories A1 and A2 is based primarily on the level of organisation of the drifting behavior.
Sequences whose drift bands follow smooth (albeit curved) tracks with little deviation are classified as A1.
On the other hand, A2 sequences show frequent interruptions, both in terms of short-duration nulls as well as rapid, but temporary, deviations in the drift rate.
These interruptions are observed to last between $\sim$ 5-10 pulses, but afterwards the drift bands reappear at the rotation phase that one would expect if the interruptions had not occurred.
These are further discussed in \S\ref{sec:P2P3DR}.

Mode B sequences are relatively short (up to 30 pulsar rotations in duration), and have small $P_3$ values ($\sim$ 10\,$P$) and large drift rates ($\sim$ $-1.2^\circ/P$).
Longer sequences ($\gtrsim 10$ rotations) are found to occur most commonly nearby to Mode A sequences; these are classified as B1.
Shorter sequences ($\lesssim 10$ rotations) may be found anywhere, even in the midst of long, otherwise uninterrupted null sequences.
As with Mode A sequences, those sequences which are too short in duration to admit more than one or two drift bands may still be positively identified by their drift rate.

@@

\begin{deluxetable*}{lcccl}
\tabletypesize{\scriptsize}
\tablewidth{0pt}
\tablecaption{Comparison between properties of several pulsars with slowly changing drift rates \label{tab:comparison}}
\tablehead{
\colhead{Pulsar} & \colhead{Nulling} & \colhead{$P_3$} & \colhead{Relaxation} & \colhead{References} \\[-8pt]
& \colhead{fraction (\%)} & \colhead{($P$)} & time, $\tau_r$ ($P$) &
}
\startdata
\psr{} & 77 & (A) 20-55 & $\gtrsim 100$ & This work \\
& & (B) $\sim10$ & & \\
B0031$-$07 & 45 & (A) $\sim12$ & $\gtrsim 50$ & \citet{Vivekanand1996,Joshi2000}, \\
& & (B) $\sim7$ & & \citep{McSweeney2017,McSweeney2019a} \\
& & (C) $\sim4$ & & \\
B0809$+$74 & 1.4 & $\sim$10.8 & $\sim$10-20 & \citet{Lyne1983,VanLeeuwen2002,VanLeeuwen2003} \\
B0826$-$34 & (weak mode) 70  & Chaotic & - & \citet{Esamdin2005,Esamdin2012,VanLeeuwen2012} \\
B0818$-$13 & 1 & 4.7 & Oscillatory & \citet{Lyne1983} \\
B0943$+$10 & $<1$ & 1.867 & $\sim$4000 & \citet{Deshpande2001,Rankin2003,Bilous2018} \\
%B1944$+$17 & $\sim$60-65 & $\sim[14,6]$ &  & \citet{Deich1986,Kloumann2010}
B2016$+$28 & $\sim1$ & $\sim3$-$12$ & - & \citet{Taylor1975,Naidu2017}
\enddata
\end{deluxetable*}

>>>>

We report the independent discovery of PSR \psr{} with the Murchison Widefield Array (MWA) in the ongoing Southern-sky MWA Rapid Two-meter (SMART) pulsar survey.
\psr{} has a period of $\sim1.306$ s, a dispersion measure (DM) of $\sim\psrdm\,$\dmunit, and a nulling fraction of $\sim$~\nullfraction{}.
This pulsar highlights the advantages of the survey's long dwell times ($\sim80\,$min), which, when fully searched, will be sensitive to the expected population of similarly bright, intermittent pulsars with long nulls.
A single-pulse analysis in the MWA's 140-170 MHz band also reveals a complex sub-pulse drifting behavior, including both rapid changes of the drift rate characteristic of mode switching pulsars, as well as a slow, consistent evolution of the drift rate within modes. In some longer drift sequences, interruptions in the otherwise smooth drift rate evolution occur preferentially at a particular phase, typically lasting a few pulses.
These properties make this pulsar an ideal test bed for prevailing models of drifting behavior such as the carousel model.

@@

Each observation was individually calibrated using {\bf observations of 3C444 taken within $\sim2$ hours of the respective target observations.
The calibration solutions were obtained using} the Real Time System software \citep{Mitchell2008}.
After calibration, the voltages were used to form a tied-array beam in the direction of the pulsar (the localization procedure is discussed in \S\ref{sec:localization}), using the bespoke beamforming software described in \citet{Ord2019}.
{\bf The MWA is located in a very radio quiet location, and no significant radio frequency interference was found in the beamformed data.}
After determining the best dispersion measure (DM) and period (see \S\ref{sec:period}), de-dispersion and folding were performed with {\tt DSPSR} \citep{VanStraten2011b} and {\tt PSRCHIVE} \citep{Hotan2004} to form single pulse archives, from which Stokes I pulse stacks were created.
Profiles of each detection, as well as an integrated profile incorporating the data from all four observations, is shown in Fig \ref{fig:profiles}.
Subsequent analysis of the pulse stacks is described in \S\ref{sec:analysis}.
\begin{figure}[!th]
    \centering
    \includegraphics[width=\linewidth]{profiles.png}
    \caption{{\bf Peak-normalised mean profiles (black) of four of the observations}, and the total mean profile (top panel) including all four observations, with a time resolution of $\sim1.3\,$ms. {\bf The two observations on MJD 59002 are not included here as they consist almost entirely of noise.} The profiles show a barely resolved double peak structure. {\bf The gray lines show the profiles made up only of pulses classified as nulls.}}
    \label{fig:profiles}
\end{figure}

@@

Follow-up of the initial MWA detection with the full 80 minutes, as well as in other archival MWA observations, revealed that the pulsar has a very large nulling fraction ($\sim$~\nullfraction{})\footnote{The nulling fraction estimate comes only from analysis of the MWA observations.}, with some nulling sequences {\bf (excluding occasional intermittent pulses) potentially exceeding $20\,$min (i.e. the length of the first observation on MJD 59002, which contained \textit{only} a few intermittent pulses, as shown in Fig. \ref{fig:maxima})}.
A full analysis of the nulling properties is presented in Section \S\ref{sec:nulling} below.
Despite the fact that \psr{} was serendipitously discovered in the first ten minutes, it may well be the harbinger for a population of relatively bright, intermittent pulsars which the SMART survey, with its $80$-minute dwell times, is well-suited to detect.
A similar strategy is employed in the Low Frequency Array (LOFAR) Tied-Array All-Sky Survey \citep[LOTAAS][]{Sanidas2019} and the low latitude ($|b| \le 5^\circ$) segment of the High Time Resolution Universe survey \citep[HTRU][]{Keith2010}, which use dwell times of approximately one hour.

@@

The discovery observation (Obs ID 1226062160) was taken when the MWA was in its \textit{compact} configuration \citep[see][]{Wayth2018}, giving an effective tied-array beam size of $\sim23^\prime$ at the central frequency of $155\,$MHz.
As per the SMART survey design, these beams are tiled across the field of view, each of which is searched for periodic candidates.
The compact configuration includes many redundant baselines, and the tied-array beam for the compact configuration is complex.
Consequently, the pulsar was detected in multiple adjacent beams, including a boresight beam that yielded the highest signal-to-noise, and several nearby grating lobes.
The ability to exploit grating lobe detections for fast-tracking candidate confirmation and localization will be discussed in the survey description paper \citetext{Bhat et al. in prep}.

The other {\bf five observations} were taken when the MWA was in its extended configuration, giving a tied-array beam size of $\sim3^\prime$ at $155\,$MHz.
Localization within {\bf the four observations in which the pulsar was bright enough} followed the identical procedure as used in \citet{Swainston2021}, i.e.\ gridding the candidate positions with overlapping tied-array beams and using signal-to-noise and the estimated beam shape to produce a likelihood map of the source position.
It is known that the ionosphere is capable of shifting the apparent position of sources by an appreciable fraction ($\sim10$-$30\%$) of the tied-array beam, and that these offsets are not necessarily corrected by the calibration process when the calibration observations are separated from the target observation, both spatially and temporally \citetext{Swainston et al., submitted}.
Nevertheless, the localization afforded by the extended array observations was sufficiently precise (a few arcminutes) to warrant follow-up with a high-resolution imaging telescope.
Accordingly, images made with Band 3 ($300$-$500\,$MHz) data from the upgraded Giant Metrewave Radio Telescope \citep[uGMRT;][]{Gupta2017}, obtained under Director's Discretionary Time (DDTC185), enabled us to confirm the pulsar's position to an accuracy of $\sim1^\prime$.
The right ascension and declination of the final best position, consistent with both the uGMRT imaging and the MWA extended configuration observations, are 00h26m37s and $-19^\circ$56'28''.

@@

Lacking a timing solution, we used PSRCHIVE's \texttt{pdmp} routine to determine the best period ($P$) and DM for each observation independently. {\bf This routine performs a brute-force grid search in both period and DM parameter space, and returns the period and DM that yields the highest profile signal-to-noise}. The weighted averages {\bf of both period and DM (respectively)} over the four observations were calculated, {\bf with the uncertainties reported by \texttt{pdmp} being} used to weight the individual measurements.
The values thus derived were $P = 1.306150 \pm 0.000005\,$s and $\text{DM} = \psrdm \pm \psrdmerr\,$\dmunit.
This simple estimate does not take the period derivative into account (in fact, it implicitly assumes $\dot{P} = 0$), and we note that a sufficiently large $\dot{P}$ would cause the period to drift significantly over the time period spanned by our four MWA observations ($\sim1.5\,$yr).
However, the pulse stacks (e.g.\ Fig. \ref{fig:pulsestacks}) folded on the above period showed no detectable slope, which we estimate would be discernible if the pulse window changed by more than $20\,$ms over the course of the observation.
Given this, we place a conservative upper limit of $|\dot{P}| \le 10^{-13}$\,s\,s$^{-1}$.

@@

\begin{figure*}[t]
    \includegraphics[width=\textwidth]{maxima_plot.png}
    \caption{Peak flux density within the pulse window as a function of time for six MWA observations (MJDs given along the $y$-axis). {\bf The shaded gray regions show detected subpulses that are included in the nulling fraction estimation, as explained in the text.} The peak fluxes were measured after smoothing the time series with a Gaussian filter ($\sim2\,$ms wide, the approximate width of a sub-pulse) to suppress the noise contribution and accentuate the contrast between nulling and burst sequences. {\bf Note that the low-level peaks in the $\sim70$-$80\,$mins time range of the MJD 58434 and in the second MJD 59002 observation} are single pulses faintly visible in the pulse stack, but which are faint because the pulsar was moving out of the MWA's primary beam.}
    \label{fig:maxima}
\end{figure*}

To estimate the nulling fraction, we first identified which pulses contained sub-pulses within the pulse window (this is a semi-automated procedure described in \S\ref{sec:drifting_analysis}).
Any pulse without a sub-pulse is then counted as a null, even if it is nested within a burst sequence.
{\bf Under this definition, we report a nulling fraction of $\sim$~\nullfraction{}.}

@@

The error on this estimate, however, will be dominated by the small number of burst/null sequences captured in these six observations.
Therefore, the true nulling fraction may ultimately prove to be much lower or higher than the $\sim$~\nullfraction{} quoted here.
Ongoing follow-up observations (and a more complete search through archival MWA data) will allow us to place much stronger constraints on the nulling fraction, and will be reported in a future publication.

@@

The distinction between the subcategories A1 and A2 is based primarily on the level of organisation of the drifting behavior.
Sequences whose drift bands follow smooth (albeit curved) tracks with little deviation are classified as A1.
On the other hand, A2 sequences show frequent interruptions, both in terms of short-duration nulls as well as rapid, but temporary, deviations in the drift rate.
These interruptions are observed to last between $\sim$~5-10 pulses, but afterwards the drift bands reappear at the rotation phase that one would expect if the interruptions had not occurred.
These are further discussed in \S\ref{sec:P2P3DR}.

Mode B sequences are relatively short (up to 30 pulsar rotations in duration), and have small $P_3$ values ($\sim10\,P$) and large drift rates ($\sim-1.2^\circ/P$).
Longer sequences ($\gtrsim 10$ rotations) are found to occur most commonly nearby to Mode A sequences; these are classified as B1.
Shorter sequences ($\lesssim 10$ rotations) may be found anywhere, even in the midst of long, otherwise uninterrupted null sequences.
As with Mode A sequences, those sequences which are too short in duration to admit more than one or two drift bands may still be positively identified by their drift rate.

@@

\begin{deluxetable*}{lcccl}
\tabletypesize{\scriptsize}
\tablewidth{0pt}
\tablecaption{Comparison between properties of several pulsars with slowly changing drift rates \label{tab:comparison}}
\tablehead{
\colhead{Pulsar} & \colhead{Nulling} & \colhead{$P_3$} & \colhead{Relaxation} & \colhead{References} \\[-8pt]
& \colhead{fraction (\%)} & \colhead{($P$)} & time, $\tau_r$ ($P$) &
}
\startdata
\psr{} & 77 & (A) 20-55 & $\gtrsim 100$ & This work \\
& & (B) $\sim10$ & & \\
B0031$-$07 & 45 & (A) $\sim12$ & $\gtrsim 50$ & \citet{Vivekanand1996,Joshi2000}, \\
& & (B) $\sim7$ & & \citep{McSweeney2017,McSweeney2019a} \\
& & (C) $\sim4$ & & \\
B0809$+$74 & 1.4 & $\sim10.8$ & $\sim$10-20 & \citet{Lyne1983,VanLeeuwen2002,VanLeeuwen2003} \\
B0826$-$34 & (weak mode) 70  & Chaotic & - & \citet{Esamdin2005,Esamdin2012,VanLeeuwen2012} \\
B0818$-$13 & 1 & 4.7 & Oscillatory & \citet{Lyne1983} \\
B0943$+$10 & $<1$ & 1.867 & $\sim4000$ & \citet{Deshpande2001,Rankin2003,Bilous2018} \\
%B1944$+$17 & $\sim$60-65 & $\sim[14,6]$ &  & \citet{Deich1986,Kloumann2010}
B2016$+$28 & $\sim1$ & $\sim3$-$12$ & - & \citet{Taylor1975,Naidu2017}
\enddata
\end{deluxetable*}
